---
title: "Modeling"
author: "Delila Medlin"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("drought_functions.R")
# source("C:/Users/delil/Desktop/Fall 2024/Research 2024/Data/Preliminary_modeling/getUSDMCountyData.R")
```

Loading in the pdsi data
```{r}
pdsi <- rast("agg_met_pdsi_1979_CurrentYear_CONUS.nc")
```
Pima County, AZ
```{r}
set.seed(1)
drought.az.pima <- read.csv("C:/Users/delil/Desktop/Fall 2024/Research 2024/Data/Preliminary_modeling/AZ_Counties/USDM-04019.csv")
pima <- clean.county.data("Arizona", "Pima", pdsi, drought.az.pima, FALSE)
pimaxy <- clean.county.data("Arizona", "Pima", pdsi, drought.az.pima, TRUE)


# model with single training and testing set
pima.train <- pima %>% sample_frac(0.80)
pima.test <- anti_join(pima, pima.train, by = 'id')

pima.rf.fit <- randomForest(WeightedAverage ~ ., 
             data = pima.train, 
             importance = TRUE)
summary(pima.rf.fit)
pima.preds <- predict(pima.rf.fit, pima.test)
RMSE(pima.test$WeightedAverage, pima.preds)

# modeling using cross validation
ctrl <- trainControl(method = "cv", number = 10)
pima.cv.rf.fit <- train(WeightedAverage ~ ., 
                   data = pima, 
                   method = "rf", 
                   trControl = ctrl)

print(pima.cv.rf.fit)


quick.rf(pima)
```
Lewis County, WV
```{r}
set.seed(1)
drought.lewis <- read.csv("USDM-54041-Lewis-WV.csv")
lewis <- clean.county.data("West Virginia", "Lewis", pdsi, drought.lewis, TRUE)

# model with single training and testing set
lewis.train <- lewis %>% sample_frac(0.80)
lewis.test <- anti_join(lewis, lewis.train, by = 'id')

lewis.rf.fit <- randomForest(WeightedAverage ~ ., 
             data = lewis.train, 
             importance = TRUE)

lewis.preds <- predict(lewis.rf.fit, lewis.test)
RMSE(lewis.test$WeightedAverage, lewis.preds) # 0.3089867

# put this exact modeling structure into a function
quick.rf(lewis)

```
Cascade County, MT
```{r}
drought.cascade <- read.csv("USDM-30013-Cascade-MT.csv")
cascade <- clean.county.data("Montana", "Cascade", pdsi, drought.cascade, FALSE)

results <- quick.rf(cascade)


```
Grady County, OK
```{r}
drought.grady <- read.csv("USDM-40051-Grady-OK.csv")
grady <- clean.county.data("Oklahoma", "Grady", pdsi, drought.grady, FALSE)

quick.rf(grady)
```
Hennepin County, MN
```{r}
drought.henn <- read.csv("USDM-27053-Hennepin-MN.csv")
henn <- clean.county.data("Minnesota", "Hennepin", pdsi, drought.henn, FALSE)

quick.rf(henn)
```

Testing Arizona as a whole state: Fips Codes 
4001-Apache
4003-Cochise
4005-Cocoinino
4007-Gila
4009-Graham
4011-Greenlee
4012-La Paz
4013-Maricopa
4015-Mohave
4017-Navajo
4019-Pima
4021-Pinal
4023-Santa Cruz
4025-Yavapai

What I worked on shortly after Meeting on 1/22 as a brute fore attempt at getting all of the data for AZ.  
```{r}
set.seed(1)

# initialize list with all of the county fips codes
az.fips <- c("04001", "04003", "04005", "04007", "04009", 
             "04011", "04012", "04013", "04015", "04017", 
             "04019", "04021", "04023", "04025", "04027")
# Initialize a list with all of the names of arizona counties
az.counties <- c("Apache", "Cochise", "Coconino", "Gila", 
            "Graham", "Greenlee", "La Paz", "Maricopa",
            "Mohave", "Navajo", "Pima", "Pinal", 
            "Santa Cruz", "Yavapai", "Yuma")
# create df to store the data from AZ
AZ.data <- data.frame(County = character(), 
                      Actual = numeric(),
                      Predicted = numeric())

# loop through each county 
for (index in seq_along(az.counties)){
  county.name <- az.counties[index]
  fips <- az.fips[index]
  
  # get the file name 
  file.name <- paste0("C:/Users/delil/Desktop/Fall 2024/Research 2024/Data/Preliminary_modeling/AZ_Counties/USDM-", fips, ".csv")
  
  # read in the data
  drought.data <- read.csv(file.name)
  
  # processing 
  clean.data <- clean.county.data("Arizona", county.name, pdsi, drought.data, FALSE)  
  
  # save the cleaned data for documentation
  filepath <- paste0("C:/Users/delil/Desktop/Fall 2024/Research 2024/Data/CleanedAzData/cleaned", county.name, ".RData" )
  save(clean.data, file = filepath)
  
  
  
  # process and save 
  results <- quick.rf(clean.data)
  
  # store the county results with the county name 
  county.results <- data.frame(County = county.name,
                               Actual = results$test,
                               Predicted = results$predictions)
  # bind the results to the big dataset
  AZ.data <- rbind(county.results, AZ.data)
}
#save(AZ.data, file = "AZpreds.RData")
#head(AZ.data)
```
Actual versus predicted to see their relationship.
```{r}

# 'interactive' aspect from chatgpt, looks cool but not super useful for abstract
library(plotly)

p <- ggplot(AZ.data, aes(x = actual, y = predicted, color = county)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  labs(
    title = "Interactive Predicted vs. Actual Values by County",
    x = "Actual",
    y = "Predicted"
  ) +
  theme_minimal()

ggplotly(p)

```
A biplot 
```{r}
ggplot(AZ.data, aes(x = actual, y = predicted, color = county)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  labs(
    title = "USDM Prediction Trends by County in Arizona",
    x = "Actual USDM",
    y = "Predicted USDM"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

```
colorblind in r will help with colors
change opaqueness in the lines 
bold axes
```{r}
# Using viridis
library(viridis)
ggplot(AZ.data, aes(x = Actual, y = Predicted, color = County)) +
  geom_smooth(method = "lm", se = FALSE, alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, color = "black", alpha = 0.3) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_viridis_d() +  # 'd' for discrete colors
  labs(
    title = "USDM Prediction Trends by County in Arizona",
    x = "Actual USDM",
    y = "Predicted USDM"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    axis.line = element_line(linewidth = 1)
  )

# OR using RColorBrewer
library(RColorBrewer)
# Replace scale_color_viridis_d() with:
scale_color_brewer(palette = "Set2")  # Set2 is colorblind-friendly

# OR using colorspace
library(colorspace)
# Replace scale_color_viridis_d() with:
scale_color_discrete_qualitative(palette = "Colorblind")
```
Boxplot of prediction errors 
```{r}
AZ.data$error <- (AZ.data$predicted - AZ.data$actual

ggplot(AZ.data, aes(x = county, y = error, fill = county)) +
  geom_boxplot(alpha = 0.6) +
  labs(
    title = "Prediction Errors by County",
    x = "County",
    y = "Error (Predicted - Actual)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Modeling to try- averaging for whole county 
pivoting longer-keeping xy coordinates
plotly county choropleth masp 

RF modeling with xy coordinates
Averaging across all dates for the cell- not sure if this is the greatest approach 
```{r}
set.seed(1)

# initialize list with all of the county fips codes
az.fips <- c("04001", "04003", "04005", "04007", "04009", 
             "04011", "04012", "04013", "04015", "04017", 
             "04019", "04021", "04023", "04025", "04027")
# Initialize a list with all of the names of arizona counties
az.counties <- c("Apache", "Cochise", "Coconino", "Gila", 
            "Graham", "Greenlee", "La Paz", "Maricopa",
            "Mohave", "Navajo", "Pima", "Pinal", 
            "Santa Cruz", "Yavapai", "Yuma")

az.data.2 <- data.frame()

# loop through each county 
for (index in seq_along(az.counties)){
  county.name <- az.counties[index]
  fips <- az.fips[index]
  
  # get the file name 
  file.name <- paste0("C:/Users/delil/Desktop/Fall 2024/Research 2024/Data/Preliminary_modeling/AZ_Counties/USDM-", fips, ".csv")
  
  # read in the data
  drought.data <- read.csv(file.name)
  
  # processing 
  clean.data <- clean.county.data("Arizona", county.name, pdsi, drought.data, TRUE)  
  
  #############
  # grouping PDSI by the date for the whole county 
  clean.data <- clean.data %>% group_by(Date) %>% 
    summarise(PDSI_Avg = mean(PDSI, na.rm = TRUE), 
              x_Avg = mean(x, na.rm = TRUE), 
              y_Avg = mean(y, na.rm = TRUE), 
              USDM_Avg = mean(USDM_Avg, na.rm = TRUE))
  
  clean.data <- clean.data %>% mutate(County = county.name) %>% 
    select( c("USDM_Avg", "PDSI_Avg", "County", "x_Avg", "y_Avg"))

  
  az.data.2 <- rbind(az.data.2, clean.data)
  
}
# both PDSI has a range of -4:7 and USDM has a range of 0:4, which is not helpful for predicting extrema (16688 rows)

# split into training and testing 
train <- az.data.2 %>% sample_frac(0.80)
test <- anti_join(az.data.2 , train)

# build model
rf.fit <-  randomForest(USDM_Avg ~ PDSI_Avg + x_Avg + y_Avg,
                      data = train,
                      ntree = 10,
                      importance = TRUE)

# predict and find RMSE
preds <- predict(rf.fit, test)


RMSE(preds, test$USDM_Avg) # 0.8166322 using averages without the location data
                           # 0.6989047 using averages with the location data, woah

rf.county.az <- randomForest(USDM_Avg ~ PDSI_Avg + County + x_Avg + y_Avg,
                      data = train,
                      ntree = 100,
                      importance = TRUE)

preds.county <- predict(rf.county.az, test)

RMSE(preds.county, test$USDM_Avg) # 0.7182475 with the addition of a county column! 

importance(rf.county.az) # 


```
I want to do the same thing but instead of averaging to make the set more manageable, taking a sample. This will help keep some of the extrema in the data, as averaging inherently gives a measure of center and doesn't capture the extrema as well. 
```{r}
set.seed(1)

# initialize list with all of the county fips codes
az.fips <- c("04001", "04003", "04005", "04007", "04009", 
             "04011", "04012", "04013", "04015", "04017", 
             "04019", "04021", "04023", "04025", "04027")
# Initialize a list with all of the names of arizona counties
az.counties <- c("Apache", "Cochise", "Coconino", "Gila", 
            "Graham", "Greenlee", "La Paz", "Maricopa",
            "Mohave", "Navajo", "Pima", "Pinal", 
            "Santa Cruz", "Yavapai", "Yuma")

az.data <- data.frame()

# loop through each county 
for (index in seq_along(az.counties)){
  county.name <- az.counties[index]
  fips <- az.fips[index]
  
  # get the file name 
  file.name <- paste0("C:/Users/delil/Desktop/Fall 2024/Research 2024/Data/Preliminary_modeling/AZ_Counties/USDM-", fips, ".csv")
  
  # read in the data
  drought.data <- read.csv(file.name)
  
  # processing 
  clean.data <- clean.county.data("Arizona", county.name, pdsi, drought.data, TRUE)  
  
  ########
  # sample a small amount of the data to make the set more manageable 
  #clean.data <- clean.data %>% sample_frac(0.0005) # consider using two stage sampling here instead?
  
  
  #clean.data <- clean.data %>% mutate(County = county.name) %>% 
    #select( c("USDM_Avg", "PDSI", "County", "x", "y", "Date"))

  
  az.data <- rbind(az.data, clean.data)
  
}

# binning data to tenth decimal place 
binned.az.data <- bin.lat.long(az.data, .25)
  
# both PDSI has a range of about -5 to 7, and USDM has a range of 0-4, good! (9022 rows)

# split into training and testing 
train <- binned.az.data %>% sample_frac(0.80)
test <- anti_join(binned.az.data, train)

# build model
rf.fit <-  randomForest(USDM_Avg ~ PDSI_Avg + x_Avg + y_Avg,
                      data = train,
                      ntree = 10,
                      importance = TRUE)


# predict and find RMSE
preds <- predict(rf.fit, test)

RMSE(preds, test$USDM_Avg) # 0.8166322 using averages without the location data
                           #  using averages with the location data
                           # 0.7756843 using a sample with averages
                           # 0.7177591 using binned lat/long with location 

# RMSE for each gridcell
test.RMSE <- test %>% group_by(bin.x, bin.y) %>% 
  summarise(RMSE = sqrt(mean((preds - USDM_Avg)^2)))


# plotting RMSE as a map, create a shape file 
# change to grid?
rmse.sf <- st_as_sf(test.RMSE, coords = c("bin.x", "bin.y"))
# plot the shape file 
ggplot(rmse.sf) +
  geom_tile(aes(x = X, y = Y, fill = RMSE)) +
  scale_color_viridis_c(option = "magma") 

rf.county.az <- randomForest(USDM_Avg ~ PDSI + County + x + y,
                      data = train,
                      ntree = 100,
                      importance = TRUE, 
                      coefficients = TRUE)

preds.county <- predict(rf.county.az, test)

RMSE(preds.county, test$USDM_Avg) # 0.742404 with the addition of a county column! 

importance(rf.county.az) # 


```

Testing
```{r}
coconino.drought <- read.csv("C:/Users/delil/Desktop/Fall 2024/Research 2024/Data/Preliminary_modeling/AZ_Counties/USDM-04005.csv")

coconino <- clean.county.data("Arizona", "Coconino", pdsi, 
                              coconino.drought, TRUE)

cocotest <- coconino %>% sample_frac(0.003)

coconinotest <- coconino %>% group_by(cell) %>% 
  summarise(PDSI_Avg = mean(PDSI, na.rm = TRUE), 
            x_Avg = mean(x, na.rm = TRUE), 
            y_Avg = mean(y, na.rm = TRUE), 
            USDM_Avg = mean(USDM_Avg, na.rm = TRUE))
```

First attempt at extracting the whole country
```{r}
countryfips <- read.csv("All Counties and FIPS codes.csv")
contUs <- filter(countryfips, ! State %in% c("AK","HI","PR"))

# loop through each county 
for (index in seq_along(az.counties)){
  county.name <- az.counties[index]
  fips <- az.fips[index]
  
  # get the file name 
  file.name <- paste0("C:/Users/delil/Desktop/Fall 2024/Research 2024/Data/Preliminary_modeling/AZ_Counties/USDM-", fips, ".csv")
  
  # read in the data
  drought.data <- read.csv(file.name)
  
  # processing 
  clean.data <- clean.county.data("Arizona", county.name, pdsi, drought.data, TRUE)  
  
  ########
  # sample a small amount of the data to make the set more manageable 
  #clean.data <- clean.data %>% sample_frac(0.0005) # consider using two stage sampling here instead?
  
  
  #clean.data <- clean.data %>% mutate(County = county.name) %>% 
    #select( c("USDM_Avg", "PDSI", "County", "x", "y", "Date"))

  
  az.data <- rbind(az.data, clean.data)
  
}
```


