---
title: "Data wrangling"
author: "Delila Medlin"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(raster)
library(sf)
library(terra)
library(tigris)
library(dplyr)
library(tidyverse)
library(tidyterra)
library(ncdf4)
```

First creating the initial raster layer for the PDSI data from the whole US, using sub-monthly gridded PDSI data from NOAA
```{r}
# add the netcdf file to a spatraster layer 
us.pdsi <- rast("agg_met_pdsi_1979_CurrentYear_CONUS.nc")
# take a look at the summary of the object
us.pdsi

# closer look at the layers
head(names(us.pdsi))
tail(names(us.pdsi))
nlyr(us.pdsi) # I think this means we have 3600 days of observations?
```
Looking at the names of the layers, they are in days but I don't quite understand the naming conventions. NOAA lists the start date as 1/1/1979.


Loading in the AZ county boundaries from the tigris package in R
```{r}
az.county <- counties(state = "Arizona", 
                      class = "sf") 
# first I'll start with just selecting pima county
pima.county <- az.county[az.county$NAME == "Pima", ]
plot(pima.county)
```

Crop the data to the az county 
```{r}
# two methods to do this- using extract and crop both from terra, I'll try both to start and then see which is better suited for this use
pima.pdsi.extracted <- terra::extract(us.pdsi, pima.county)

# then we try the crop function from terra
pima.pdsi.cropped <- terra::crop(us.pdsi, pima.county)

# quick visualization to see what's going on
plot(pima.pdsi.cropped)
```
Not as high resolution as I was expecting given the map on the website. Maybe try plotting this differently using ggplot?


#####################################################
First Look at the Living Blended Drought Atlas Data
```{r}
setwd("C:/Users/dgm239/Downloads/Research_2025/PDSI_research_2024/Data")
lbda <- rast("C:/Users/dgm239/Downloads/Research_2025/PDSI_research_2024/Data/lbda-v2_kddm_pmdi_2017.nc")


```


